---
title: "Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE, 
  warning = FALSE,
  comment = "#>"
)
```

#### *Marcus W. Beck, mbeck@tbep.org*

# Overview 

The System Wide Monitoring Program ([SWMP](https://coast.noaa.gov/digitalcoast/data/nerr.html)) was implemented by the National Estuarine Research Reserve System ([NERRS](https://coast.noaa.gov/nerrs/)) in 1995 to provide continuous monitoring data at over 140 continuous monitoring stations in 28 estuaries across the United States.  SWMPr (pronounced "swamper") is an R package for retrieving, organizing, and analyzing estuary monitoring data from SWMP. Please cite the package as follows:

*Beck MW. 2016. SWMPr: An R package for retrieving, organizing, and analyzing environmental data for estuaries.  The R Journal. 8(1):219-232. https://journal.r-project.org/archive/2016-1/beck.pdf*

# Installing the package

Install the package from CRAN as follows:

```{r eval = F, message = F}
install.packages('SWMPr')
library(SWMPr)
```

The development (unstable) version of this package can be installed from Github:

```{r eval = F, message = F}
install.packages('devtools')
library(devtools)
install_github('fawda123/SWMPr', ref = 'development')
library(SWMPr)
```

# Using the package

A manuscript describing full use of the package is available from the [R Journal](https://journal.r-project.org/archive/2016/RJ-2016-015/index.html). All source materials for the manuscript are available [here](https://github.com/fawda123/swmpr_manu).  A brief description of the available functions is provided below. See help documentation for more details on each function (e.g., `?all_params`).  All functions are grouped into respective categories that describe their use: retrieve, organize, and analyze.  Help files for functions in each category can be viewed as follows:

```{r eval = F, message = F}
help.search('retrieve', package = 'SWMPr')
help.search('organize', package = 'SWMPr')
help.search('analyze', package = 'SWMPr')
```

<h3>Retrieve</h3>
<table>
<tr><td><code>all_params</code></td><td>Retrieve any number of records starting with the most recent at a given station, all parameters. Wrapper to <code>exportAllParamsXMLNew</code> function on web services.</td></tr>
<tr><td><code>all_params_dtrng</code></td><td> Retrieve records of all parameters within a given date range for a station. Optional argument for a single parameter. Wrapper to <code>exportAllParamsDateRangeXMLNew</code>.</td></tr>
<tr><td><code>import_local</code></td><td> Import files from a local path. The files must be in a specific format, specifically those returned from the CDMO using the <a href="http://cdmo.baruch.sc.edu/aqs/zips.cfm">zip downloads</a> option for a reserve.</td></tr>
<tr><td><code>import_remote</code></td><td> Import SWMP site data from a remote independent server. These files have been downloaded from CDMO, processed using functions in this package, and uploaded to an Amazon server for quicker import into R.</td></tr>
<tr><td><code>single_param</code></td><td> Retrieve any number of records for a single parameter starting with the most recent at a given station. Wrapper to <code>exportSingleParamXMLNew</code> function on web services.</td></tr>
<tr><td><code>site_codes</code></td><td> Metadata for all stations, wrapper to <code>exportStationCodesXMLNew</code> function on web services.</td></tr>
<tr><td><code>site_codes_ind</code></td><td> Metadata for all stations at a single site, wrapper to <code>NERRFilterStationCodesXMLNew</code> function on web services.</td></tr>
</table>
<h3>Organize</h3>
<table>
<tr><td><code>cens_id</code></td><td> Creates additional columns in a swmpr object that identify observations below, within, or above the detection limit.</td></tr>
<tr><td><code>comb</code></td><td> Combines swmpr objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.</td></tr>
<tr><td><code>qaqc</code></td><td> Remove QAQC columns and remove data based on QAQC flag values for a swmpr object. Only applies if QAQC columns are present.</td></tr>
<tr><td><code>qaqcchk</code></td><td> View a summary of the number of observations in a swmpr object that are assigned to different QAQC flags used by CDMO. The output is used to inform further processing but is not used explicitly.</td></tr>
<tr><td><code>rem_reps</code></td><td> Remove replicate nutrient data that occur on the same day. The default is to average replicates.</td></tr>
<tr><td><code>setstep</code></td><td> Format data from a swmpr object to a continuous time series at a given timestep. The function is used in <code>comb</code> and can also be used with individual stations.</td></tr>
<tr><td><code>subset</code></td><td> Subset by dates and/or columns for a swmpr object. This is a method passed to the generic `subset’ function provided in the base package.</td></tr>
</table>
<h3>Analyze</h3>
<table>
<tr><td><code>aggreswmp</code></td><td> Aggregate swmpr objects for different time periods - years, quarters, months, weeks, days, or hours. Aggregation function is user-supplied but defaults to mean.</td></tr>
<tr><td><code>aggremetab</code></td><td> Aggregate metabolism data from a swmpr object. This is primarily used within <code>plot_metab</code> but may be useful for simple summaries of raw daily data.</td></tr>
<tr><td><code>ecometab</code></td><td> Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.</td></tr>
<tr><td><code>decomp</code></td><td> Decompose a swmpr time series into trend, seasonal, and residual components. This is a simple wrapper to <code>decompose</code>. Decomposition of monthly or daily trends is possible.</td></tr>
<tr><td><code>decomp_cj</code></td><td> Decompose a swmpr time series into grandmean, annual, seasonal, and events components. This is a simple wrapper to <code>decompTs</code> in the wq package. Only monthly decomposition is possible.</td></tr>
<tr><td><code>hist</code></td><td> Plot a histogram for a swmpr object.</td></tr>
<tr><td><code>lines</code></td><td> Add lines to an existing swmpr plot.</td></tr>
<tr><td><code>map_reserve</code></td><td> Create a map of all stations in a reserve using the ggmap package.</td></tr>
<tr><td><code>na.approx</code></td><td> Linearly interpolate missing data (<code>NA</code> values) in a swmpr object. The maximum gap size that is interpolated is defined as a maximum number of records with missing data.</td></tr>
<tr><td><code>overplot</code></td><td> Plot multiple SWMP time series on the same y-axis.</td></tr>
<tr><td><code>plot</code></td><td> Plot a univariate time series for a swmpr object. The parameter name must be specified.</td></tr>
<tr><td><code>plot_metab
</code></td><td> Plot ecosystem metabolism estimates after running <code>ecometab</code> on a swmpr object.</td></tr>
<tr><td><code>plot_summary</code></td><td> Create summary plots of seasonal/annual trends and anomalies for a water quality or weather parameter.</td></tr>
<tr><td><code>plot_wind</code></td><td> Create a wind rose using weather data.</td></tr>
<tr><td><code>smoother</code></td><td> Smooth swmpr objects with a moving window average. Window size and sides can be specified, passed to <code>filter</code>.</td></tr>
</table>
<h3>Miscellaneous</h3>
<table>
<tr><td><code>calcKL</code></td><td> Estimate the reaeration coefficient for air-sea gas exchange. This is only used within the <code>ecometab</code> function.</td></tr>
<tr><td><code>metab_day</code></td><td> Identify the metabolic day for each approximate 24 period in an hourly time series. This is only used within the <code>ecometab</code> function.</td></tr>
<tr><td><code>param_names</code></td><td> Returns column names as a list for the parameter type(s) (nutrients, weather, or water quality). Includes QAQC columns with ‘f_’ prefix. Used internally in other functions.</td></tr>
<tr><td><code>parser</code></td><td> Parses html returned from CDMO web services, used internally in retrieval functions.</td></tr>
<tr><td><code>swmpr</code></td><td> Creates object of swmpr class, used internally in retrieval functions.</td></tr>
<tr><td><code>time_vec</code></td><td> Converts time vectors to POSIX objects with correct time zone for a site/station, used internally in retrieval functions.</td></tr>
</table>
